---
title: "Dental Caries Trajectories"
author: "Chukwuebuka Ogwo"
date: "7/25/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

```
* Description of data:  
  - A cohort study (subjects were recruited at birth and followed for 25 years).
  - A high dimensional data (more than 50 relevant variables)
  - Dental examinations were performed at age 5, 9, 13, 17 and 23. 
  - Independent variables data were collected at every 6 months. 
  - Almost 2000 subjects were recruited at birth but only about 336 subjects were retained after 25 years (a lot of missing data).
  
* Description: A trajectory analysis of dental caries over a life course using a longitudinal data.
* Aim: To use unsupervised machine learning algorithm (KmL) to understand the changes in caries incidence from childhood to adulthood. 
* Ultimate goal: To help cities/health departments identify high risk groups and help them streamline their interventions which will save them time and money.  

```{r packages}
library("kml3d")
library("kml")
library("rgl")
library("dplyr")
library("psych")
library("tableone")
require(MASS)
```

```{r dataset}
### 1. Importing new dataset
setwd("/Users/damia/OneDrive/Documents/PhD/Datasets")
library(haven)
CariesTraj <- read_sas("/Users/damia/OneDrive/Documents/PhD/Datasets/dfs_increments.sas7bdat")
View(CariesTraj)
head(CariesTraj)
#CariesTraj2<- CariesTraj[c(1,2,4:6)]#
#CariesTraj0<- na.omit(CariesTraj)#
#CariesTraj1<- CariesTraj0[c(1,2,4:6)]#
```

## Including Plots
```{r Longitudinal data clustering }
### 2. Main analysis### Longitudinal clustering for Trajectory analysis
Ebtraj<-cld(traj=CariesTraj,
    idAll=as.character(CariesTraj$id),
    time=c(9,13,17,23),
    timeInData=c(2,4:6), maxNA=2)
 kml(Ebtraj,nbClusters = 3:6, nbRedrawing=20,toPlot="both")
```


```{r Partition selection}
 ### 3. Get and choose best partions
X11()
choice(Ebtraj, typeGraph = "png") ### To see the best partition  
plotAllCriterion(Ebtraj)          ### To check the best's cluster numbers##
print(Ebtraj)
mean.trajectories <- calculTrajMean(Ebtraj["traj"], Ebtraj['c4'][[1]]['clusters'])
print(mean.trajectories)
```

*Use case: The sudden rise and plateauing in the trajectories between age 13 and 17 suggests that btw age 13 and 17 is a high risk period.
           Currently, the Government provides millions of dollars towards dental sealant interventions from age 6 to 19. So with the findings above,
           they can channel their funding towards the higher risk age group (btw age 13 to 17) thereby saving them millions of dollars.    

```{r Data export and merging}
### 4. Export the clusters and create a new dataset that includes the clusters
CariesTraj$cluster <- getClusters(Ebtraj,4)
View (CariesTraj)
```


